---
title: "ps3 solutions"
author: "Irene Ann Iype"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## Question 1
#### a) Does territory size have a causal influence the weight of foxes? Construct a quap model to infer the total causal influence of area on weight. Does increasing the area available to each fox make it healthier (i.e., heavier)?


```{r}
library(rethinking)

data(foxes)
d <- foxes
precis( d )

library(dagitty)
dag1 <- dagitty("dag{area -> avgfood -> weight; avgfood -> groupsize ->weight}")
plot(dag1)
impliedConditionalIndependencies(dag1)

```

We look at the relationship between area and weight. From the DAG provided, the paths are:  

1) area -> avgfood -> weight (open path)

2) area -> avgfood -> groupsize -> weight (open path)

Both are causal paths with no counfounding. i.e the estimate of the effect of area on weight is not confounded by any other variable. No backdoor paths because there are no arrows entering area. Therefore, we do condition on avgfood or groupsize, as this would close one or more of the paths shown above and we want to use both to understand the total
effect of area on weight.

Standardising variables:

``` {r}
#standardise variables - mean 0, sd1
d$area <- standardize( d$area )
d$weight <- standardize( d$weight )
```

Using prior predictive simulation to show that priors make sense (code from pg95):
```{r}
set.seed(2971)
N <- 100                  
a <- rnorm( N , 0 , 0.2 )
bA <- rnorm( N , 0 , 0.5 )

plot( NULL , xlim=range(d$area) , ylim=c(-3,3) ,
      xlab="standardised area" , ylab="standardised weight" )

#minimum & maximum line 
abline( h= max(d$weight) , col=2 ) #max weight
abline( h= min(d$weight) , col=3 ) #min weight

mtext( "b ~ dnorm(0,10)" )
xbar <- mean(d$area)
for ( i in 1:N ) curve( a[i] + bA[i]*x,
                        from=min(d$area) , to=max(d$area) , add=TRUE ,
                        col=col.alpha("black",0.2) )

```

MLR model:

```{r}
m1a <- quap(
  alist(
    weight ~ dnorm( mu , sigma ) ,
    mu <- a + bA * area ,
    a ~ dnorm( 0 , 0.2 ) ,
    bA ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 )
  ) , data = d )

precis(m1a)
```
bA is 0.02 which shows there is no correlation between area and weight. No causal effect of area on weight.

#### b) Now infer the causal impact of adding food (avgfood) to a territory. Would this make foxes heavier? Which covariates do you need to adjust to estimate the total causal influence of food?

We look at the relationship between avgfood and weight. 
Paths: 

1) avgfood -> weight (open path)

2) avgfood -> groupsize -> weight (open path)

No backdoors since there are no arrows entering avgfood. The effect of food on weight can be inferred by considering both the paths above. Once again, we do not need to adjust for groupsize since we need both paths to stay open so we know the TOTAL effect of avgfood. Conditioning on groupsize would close path 2 (indirect path).

``` {r}
#standardising food
d$avgfood <- standardize(d$avgfood)

#using multiple regression like before
m1b <- quap(
  alist(
    weight ~ dnorm( mu , sigma ) ,
    mu <- a + bF * avgfood ,
    a ~ dnorm( 0 , 0.2 ) ,
    bF ~ dnorm( 0 , 0.5 ) ,
    sigma ~ dexp( 1 )
  ), data = d )

precis(m1b)
```

bF is almost 0. Once again, there is no causal impact on weight from the average food. Intuitively, we would think avgfood would impact weight, however considering the DAG provided, this result is expected since area is directly upstream of food. And since area doesn't show any correlation with weight, the same can be expected for food.

#### c) Now infer the causal impact of group size (groupsize). Which covariates do you need to adjust to make this estimate? Inspect the posterior distribution of the resulting model. What do you think explains these data? Specifically, explain the estimates of the effects of area, avgfood,and groupsize on weight. How do they make sense together? 

There are 2 paths between group size and weight: 

1) groupsize -> weight (open)

2) groupsize <- avgfood -> weight (backdoor path - must close)

The second one is a backdoor path since there is an arrow going into groupsize (from avgfood). So we need to close this backdoor -> we need to use avgfood in our model as well as groupsize

```{r}
#standardise 
d$groupsize <- standardize(d$groupsize)

#using standard priors from earlier 
m1c <- quap(
  alist(
    weight ~ dnorm( mu , sigma ) ,
    mu <- a + bG*groupsize + bF*avgfood ,
    a ~ dnorm( 0 , 0.2 ) ,
    bF ~ dnorm(0, 0.5),
    bG ~ dnorm(0, 0.5),
    sigma ~ dexp( 1 )
  ),
  data=d )

precis( m1c )
```

From the results in m1c, we can see that bG = -0.57 which indicates that group size is correlated negatively with weight (when food is held constant) while food is showing a positive correlation with weight (when group size is held constant). This makes sense since more food means more food per fox and thus increased weight while a larger group size means less food per fox (more sharing) which leads to a decrease in weight.

However, 1b showed that food doesnt have any correlation with weight and this is because its effects are being cancelled out through groupsize. The total causal effect of food is minimal since it also causes larger groups. This shows a masking effect which is also called post-treatment bias.

The causal explanation is that as food increases, the groupsize increases until the food per fox is more or less the same among all territories. Result: food has no total significant effect on the weight of foxes 


## Question 2
#### Explain the difference between model selection and model comparison. What information is lost under model selection?

Model selection has to do with choosing criteria which help in choosing model structure. A common form of model selection is to choose a model which has the best information criteria value or where every coefficient is statistically significant ('stargazing' as mentioned in the book). However, this kind of selection procedure discards the information about relative model accuracy contained in the differences among the CV/PSIS/WAIC values.

Another downside to model selection using these criterion alone is that they do not tell you anything about the causal structure of the parameters. It is totally possible for a model to be a great predictor or forecaster but be highly confounded. These models do not help us if our aim is to make some intervention and figure out what the result of these interventions might be. So it can be especially problematic when one model outperforms its alternatives to a small degree and we simply pick it purely on its IC values.

On the plus side, these criteria do account for overfitting and help in testing model implications, given a set of causal models. This leads to a better approach: Model comparison, which is a more general approach where we use multiple models to understand how different variables influence predictions. This, along with a causal model with implied conditional independencies among variables help us infer causal relationships.

## Question 3
#### Can you explain the relative differences in WAIC scores, using the fox DAG from above? 

```{r}
# Constructing the models 
data(foxes)
d <- foxes 

#standardising weights
d$weight <- standardize(d$weight)
d$area <- standardize(d$area)
d$avgfood <- standardize(d$avgfood)
d$groupsize <- standardize(d$groupsize)

#model 1: avgfood + groupsize + area

m1 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bF*avgfood + bG*groupsize + bA*area, #mlr
    a ~ dnorm(0,0.2),
    c(bF,bG,bA) ~ dnorm(0,0.5), #using priors from earlier
    sigma ~ dexp(1)
  ), data=d )

#model 2: avgfood + groupsize

m2 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bF*avgfood + bG*groupsize, #mlr
    a ~ dnorm(0,0.2),
    c(bF,bG) ~ dnorm(0,0.5), #using priors from earlier
    sigma ~ dexp(1)
  ), data=d )

#model 3: avgfood + area
m3 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bF*avgfood + bA*area, #mlr
    a ~ dnorm(0,0.2),
    c(bF,bA) ~ dnorm(0,0.5), #using priors from earlier
    sigma ~ dexp(1)
  ), data=d )

#model 4: avgfood 
m4 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bF*avgfood, #mlr
    a ~ dnorm(0,0.2),
    bF ~ dnorm(0,0.5), #using priors from earlier
    sigma ~ dexp(1)
  ), data=d )

#model 5: area
m5 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + bA*area, #mlr
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5), #using priors from earlier
    sigma ~ dexp(1)
  ), data=d )

compare(m1,m2,m3,m4,m5)
```

Columns from left to right are: WAIC, standard error (SE) of WAIC, difference of each WAIC from the best model, standard error (dSE) of this difference, prediction penalty (pWAIC), and finally the Akaike weight

The best models based on the WAIC values are m1 & m2. Their WAIC values are also very close to each other. The dWAIC column is the difference between each model’s WAIC and the best WAIC in the set. To figure out if these models can be easily distinguished by their out of sample accuracy, we need to consider the error in the WAIC estimates. For this we use SE & dSE. 

```{r}
plot( compare( m1 , m2 , m3, m4, m5 ) )
```
The differences of in the WAIC scores all fall well within the 99% intervals of the differences

m1 and m2 are very close to each other as shows in the plot for their out of sample deviance. This makes sense since both of these models have groupsize in it which means we can end up with the same results 
when including the other parameters. This has to do with the DAG structure from the previous question. The first model includes food, area and groupsize while the second only involves 
food and groupsize. However, food has a direct causal relationship to area which is routed entirely through food. So the affect of area while adjusting for groupsize is the same as the effect of food while adjusting for groupsize. Including groupsize essentially closes the backdoor path so both models have similar predictive power.

m4,5,3 also perform similarly. This is because they all do not include groupsize. So for model 5(only area) would be very similar to model 4(only food) since area has to pass through food which leads to the same effect.
Model 3 includes both area and food, but again because of the way area is routed through food, this also leads to a very similar result as the other two. However we can also note that model 3 performs the worst out of 4,5,6.
The total causal influence of area or food alone is about 0 as shown in Q1a and b. 


